---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# monoClust

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/vinhtantran/monoClust.svg?branch=master)](https://travis-ci.org/vinhtantran/monoClust)
[![CRAN status](https://www.r-pkg.org/badges/version/monoClust)](https://CRAN.R-project.org/package=monoClust)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/monoClust)](https://cran.r-project.org/package=monoClust)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![R-CMD-check](https://github.com/vinhtantran/monoClust/workflows/R-CMD-check/badge.svg)](https://github.com/vinhtantran/monoClust/actions)
[![Codecov test coverage](https://codecov.io/gh/vinhtantran/monoClust/branch/master/graph/badge.svg)](https://codecov.io/gh/vinhtantran/monoClust?branch=master)
<!-- badges: end -->

Implementation of the Monothetic Clustering algorithm on continuous data set. A lot of extensions are included in the package, including applying Monothetic clustering on data set with circular variables, visualizations with the results, and permutation and cross-validation based tests to support the decision on the number of clusters.

The package performs Monothetic clustering on numerical variables only data. The extended features include clustering on one circular variable, limiting the sets of splitting variables. There are also S3 functions such as print and plot that works for MonoClust output object.

## Installation

You can install the released version of monoClust from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("monoClust")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("vinhtantran/monoClust")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(monoClust)
library(cluster)
data(ruspini)
ruspini4sol <- MonoClust(ruspini, nclusters = 4)
ruspini4sol
```

Plotting MonoClust tree:

```{r cars}
plot(ruspini4sol)
```
